Practical Ethics for Computational Semantics\\ \large Philosophy of Distributional Semantics

### 2017-06-05 pds/Paper/paper_pds.tex
dis the end

\documentclass{article}
\title{Practical Ethics for Computational Semantics\\
\large Philosophy of Distributional Semantics}
\date{}
\author{Silvan Hungerb{\"u}hler}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{biblatex}
\addbibresource{pds_references.bib}
\usepackage{csquotes}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of Distributional Semantics (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbors the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants draw on one particular attempt to provide general-purpose guidelines for ML - the \emph{Principles for Accountable Algorithms} to suggest guidelines and best practices for computational methods in DS.
As the phenomena of interest in DS differs in important aspects from other applications of ML, notably also natural language processing,
this involves assessing the applicability of general purpose ML guidelines to DS. Our main finding is that most principles carry over smoothly, except for issues concerning responsibility and algorithmic fairness.
\end{abstract}
\section{Introduction}
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications \cite{algorithms2016}.
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

@@

As recently as 2016, however, a discussion has opened up in a hitherto completely undiscussed domain. ML techniques used for the computer-based study of language meaning - a field known as \emph{computational distributional semantics} - has come under scrutiny.
Various studies have demonstrated that these methods can  yield results which are in important ways biased against particular groups as, for example, women \cite{bolukbasi2016man}\cite{wagner2015s}.
This form of machine bias in word meaning can affect people in discriminatory ways. It is quite conceivable that in the near future many applications, for example automated customer services, will operate with automatically generated representations of natural language meaning. If semantic bias translates into discriminatory decisions being taken, then there is a problem.

@@

The present contribution works by scanning previous efforts made in a similar vein, identifying useful parts therein and applying them to ethical questions surrounding DS. We will make no foundational efforts to sustain the validity of the principles. They are presupposed. Our analysis merely serves to discuss the usefulness of the principles in particular domain, and assert them where due.
Currently, we want to focus on aspects of discrimination and bias in the outcome of DS. This is for two reasons: For one, these are the least explored ethical issues for DS. Furthermore, some of the ideas, guidelines and principles concerning privacy, abuse of otherwise legitimate technology ot  or economic impact are directly transferable from the general ML debate to any sub-field that similarly employs big data and algorithms. Computational DS is no exception. Debates concerning privacy etc. can follow the same lines as the discussion to establish general purpose ethical principles for ML. The issue of discrimination and bias, however, has a special relation to DS because the field occupies a unique position of computational language research. The central concern of DS, as its name gives away, is the concept of meaning. As we will argue presently, this relates it in a special way to many sensitive topics in the general ML discussion. Put succinctly, it sits at the intersection of concerns pertaining to data selection, validity or questionableness of computational methods, interpretation of its results and the use it is given in applications.

@@

In many cases of undesirable effects that come with these technologies, legislation is behind in adequately protecting people negatively affected by it.
Given this state of affairs, it is important to hold the people who develop or implement such algorithmic systems accountable to the public. The FATML concisely write in their PAA\cite{principles}: 
\begin{quote}
Algorithms and the data that drive them are designed and created by people -- There is always a human ultimately responsible for decisions made or informed by an algorithm. ''The algorithm did it'' is not an acceptable excuse if algorithmic systems make mistakes or have undesired consequences, including from machine-learning processes.
\end{quote}

@@

Concerning the question of fairness, consider a concrete scenario where the problem arises: A recent study has found, for example, that a model of lexical word meaning trained on Google News articles displays significant bias along binary gender dimensions \cite{bolukbasi2016man}. Despite the articles in the corpus being written by professional journalists, the word \emph{woman} bears the same semantic relation to \emph{homemaker} in the meaning space as the word \emph{man} to \emph{computer scientist}. A search engine could use semantic information available in such a model, with bias and all, to optimize the relevance of their search results. Imagine an employer querying Google for ''Computer Scientist Kabul'' to find professionals in the area and Google deems LinkedIn profiles with male-associated names more relevant to the search request. Consequently, male computer scientists' profiles are ranked higher in the results and thus more visible. This would be a clear-cut case of discrimination and disadvantage based on gender, brought about directly by recommendation mechanism which utilizes DS. These are not mere hypotheticals though, but already very concrete issues. It is has been reported that when one queries Google with the name of a person, then Google is more likely to show advertisement related to that person's criminal record if the name is associated with black people \cite{sweeney2013discrimination}.

This raises a host of questions concerning the provenance of the bias in the model, the adequate response by the responsible distributional semanticist and, not to mention, the nature of meaning.
Does one have to live with the fact that the meanings discovered in some corpus systematically contain unfavorable connotations for some group of people? After all, the corpus was produced by a particular society, so it should be expected that it mirror that society's prejudices. What position should one assume towards these results? Is it fair to claim that the the model picked up what it is supposed to pick up and blame society for systematically biased meanings? Or does whoever works with the model the obligation to rid it of biased meaning representations? Such questions gain concrete importance if the semantic model is further used for some interaction within that same society that taught it - via corpus - what it means to be a \emph{woman}, for example. 
What if using the model results in re-enforcement of the structures that lead to the biased corpus in the first place? 
Who is to blame for this unfortunate outcome? The recommendation systems designers might simply point at the corpus in which they found their word meanings. After all, the corpus accurately reflects the fact that there are more men than women who work in computer science \cite{womenincs}. So the biased outcome of the search mirrors societal structure, or so would the recommender's excuse might sound.

@@

The lesson of this particular successfull debiasing endeavor is straightforward. Performance is no general excuse for designers of systems employing DS to work with meanings that reflect societal bias. Efforts can be made to process the DS model after its original training to rid it of discriminatory disparities. In general domain ML there is already a sizable body of literature of finding an removing so-called \emph{disparate impact} - unintentional discrimination (cf. \cite{feldman2015certifying} \cite{dwork2012fairness}). 
Similar collections of standardized techniques should be developed to debias semantic spaces. Doing this is especially important, of course, if the DS model to be used for some outside application. 
Perhaps a complement to the attempts to debiase embeddings, then, are the studies whose stated goal it is to uncover biases in corpora (cf. \cite{wagner2015s} \cite{herbelot2012distributional}). Findings that indicate clear instances of bias in widely used corpora can then be used as an indication where  debiasing should have to take place before the DS is put to further use.
Thus, it appears that a two-fold effort is required in to prevent damage from happening in the employment of algorithmically generated meanings. On the one hand, one will constantly have to scan the semantic spaces gathered from corpora for undesired connotations or other semantic relationships within it; on the other, once such biases are certified, efforts need to be undertaken, so that systems using this semantic information do not systematically disadvantage people.

>>>>

\documentclass{article}
\title{Practical Ethics for Computational Semantics\\
\large Philosophy of Distributional Semantics}
\date{}
\author{Silvan Hungerb{\"u}hler}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{biblatex}
\addbibresource{pds_references.bib}
\usepackage{csquotes}
\fontsize{10}{12}
\usepackage[urw-garamond]{mathdesign}
\usepackage[margin=5cm]{geometry}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of Distributional Semantics (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbors the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants draw on one particular attempt to provide general-purpose guidelines for ML - the \emph{Principles for Accountable Algorithms} to suggest guidelines and best practices for computational methods in DS.
As the phenomena of interest in DS differs in important aspects from other applications of ML, notably also natural language processing,
this involves assessing the applicability of general purpose ML guidelines to DS. Our main finding is that most principles carry over smoothly, except for issues concerning responsibility and algorithmic fairness.
\end{abstract}
\section{Introduction}
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications \cite{algorithms2016}.
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

@@

As recently as 2016, however, discussion has opened up in a hitherto completely undiscussed domain. ML techniques used for the computer-based study of language meaning - a field known as \emph{computational distributional semantics} - has come under scrutiny.
Various studies have demonstrated that these methods can  yield results which are in important ways biased against particular groups as, for example, women \cite{bolukbasi2016man}\cite{wagner2015s}.
This form of machine bias in word meaning can affect people in discriminatory ways. It is quite conceivable that ian the near future many applications, for example automated customer services, will operate with automatically generated representations of natural language meaning. If semantic bias translates into discriminatory decisions being taken, then there is a problem.

@@

The present contribution works by scanning previous efforts made in a similar vein, identifying useful parts therein and applying them to ethical questions surrounding DS. We will make no foundational efforts to sustain the validity of the principles. They are presupposed. Our analysis merely serves to discuss the usefulness of the principles in particular domain, and assert them where due.
Currently, we want to focus on aspects of discrimination and bias in the outcome of DS. These are the least explored ethical issues for DS. Furthermore, some of the ideas, guidelines and principles concerning privacy, abuse of otherwise legitimate technology or economic impact are directly transferable from the general ML debate to any sub-field that similarly employs big data and algorithms. Computational DS is no exception. Debates concerning privacy etc. can follow the same general purpose ethical principles for ML. The issue of discrimination and bias, however, has a special relation to DS because the field occupies a unique position of computational language research. The central concern of DS, as its name gives away, is the concept of meaning. As we will argue presently, this relates it in a special way to many sensitive topics in the general ML discussion. Put succinctly, it sits at the intersection of concerns pertaining to data selection, validity or questionableness of computational methods, interpretation of its results and the use it is given in applications.

@@

\newpage In many cases of undesirable effects that come with these technologies, legislation is behind in adequately protecting people negatively affected by it.
Given this state of affairs, it is important to hold the people who develop or implement such algorithmic systems accountable to the public. The FATML concisely write in their PAA\cite{principles}: 
\begin{quote}
Algorithms and the data that drive them are designed and created by people -- There is always a human ultimately responsible for decisions made or informed by an algorithm. ''The algorithm did it'' is not an acceptable excuse if algorithmic systems make mistakes or have undesired consequences, including from machine-learning processes.
\end{quote}

@@

Concerning the question of fairness, consider a concrete scenario where the problem arises: A recent study has found that a model of lexical word meaning trained on Google News articles displays significant bias along binary gender dimensions \cite{bolukbasi2016man}. Despite the articles in the corpus being written by professional journalists, the word \emph{woman} bears the same semantic relation to \emph{homemaker} in the meaning space as the word \emph{man} to \emph{computer scientist}. A search engine could use semantic information available in such a model, with bias and all, to optimize the relevance of their search results. Imagine an employer querying Google for ''Computer Scientist Kabul'' to find professionals in the area and Google deems LinkedIn profiles with male-associated names more relevant to the search request. Consequently, male computer scientists' profiles are ranked higher in the results and thus more visible. This would be a clear-cut case of discrimination and disadvantage based on gender, brought about directly by recommendation mechanism which utilizes DS. These are not mere hypotheticals though, but already very concrete issues. It is has been reported that when one queries Google with the name of a person, then advertisement related to that person's criminal record is likelier if the name is associated with black people \cite{sweeney2013discrimination}.

This raises a host of questions concerning the provenance of the bias in the model, the adequate response by the responsible distributional semanticist and, not to mention, the nature of meaning.
Is it necessary that the meanings discovered in some corpus systematically contain unfavorable connotations for some group of people? After all, the corpus was produced by a particular society, so it should be expected that it mirror that society's prejudices. What position should one assume towards these results? Is it fair to claim that the the model picked up what it is supposed to pick up and blame society for systematically biased meanings? Or does whoever works with the model the obligation to rid it of biased meaning representations? Such questions gain concrete importance if the semantic model is further used for some interaction within that same society that taught it - via corpus - what it means to be a \emph{woman}, for example. 
What if using the model results in re-enforcement of the structures that lead to the biased corpus in the first place? 
Who is to blame for this unfortunate outcome? The recommendation systems designers might simply point at the corpus in which they found their word meanings. After all, the corpus accurately reflects the fact that there are more men than women who work in computer science \cite{womenincs}. So does the biased outcome of the search only mirror societal structure? %, or so would the recommender's excuse sound.

@@

The lesson of this particular successfull debiasing endeavor is straightforward. Performance is no general excuse for designers of systems employing DS to work with meanings that reflect societal bias. Efforts can be made to process the DS model after its original training to rid it of discriminatory disparities. In general domain ML there is already a sizable body of literature of finding an removing so-called \emph{disparate impact} - unintentional discrimination (cf. \cite{feldman2015certifying} \cite{dwork2012fairness}). 
Similar collections of standardized techniques should be developed to debias semantic spaces. Doing this is especially important, of course, if the DS model to be used for some outside application. 
Perhaps a complement to the attempts to debiase embeddings, then, are the studies whose stated goal it is to uncover biases in corpora (cf. \cite{wagner2015s} \cite{herbelot2012distributional}). Findings that indicate clear instances of bias in widely used corpora can then be used as an indication where  debiasing should have to take place before the DS is put to further use.
Thus, it appears that a two-fold effort is required: On the one hand, one will constantly have to scan the semantic spaces gathered from corpora for undesired connotations or other semantic relationships within it; on the other, once such biases are certified, efforts need to be undertaken, so that systems using this semantic information do not systematically disadvantage people.
